Modelling the emergence of whisker barrels

### 2020-01-27 paper/briefpaper/brief.tex
Changes following meeting with Stuart

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  The
assumption of pre-arranged center-points \cmnt{(an assumption shared with
  chemoaffinity theories following \citealp{sperry_chemoaffinity_1963})} is
difficult to resolve with the observation that axons arrive in the cortical
plate as an undifferentiated bundle, \emph{prior} to barreloid formation
\citep{agmon_organized_1993}.  \cmnt{But it's not that hard to resolve, if, as
  in the paper, we assume orthogonal gradients in the barreloid field.}

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. The second term on the right represents
an assumption that axon branching decreases when synaptogenesis increases, and
vice versa, i.e., \cmnt{when a connection is made, the axon branching
  decreases by one branch and when a connection is lost, one branch is added
  to the branching density}
%that axon branches are required for synaptic connections to form.
Here $\chi_i=0$ is a placeholder.

@@

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density at $t=0$ \cmnt{(less the
  number of connections that have been made).};
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Normalization
  corresponds to the biological constraint that the branching and growth of
  thalamocortical axons is backed by a finite resource provided by their
  parent cell bodies.}

@@

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity \cmnt{(and an associated
  in-group/out-group axonal identification mechanism)}, and also in terms of
cortical space (as represented by $\chi$). This requirement may account for
the absence of barrels in rodents with larger brains, such as capybara, for
which competition for space is presumably weaker
\citep{woolsey_comparative_1975}. Hence, irrespective of whether barrels are
necessary for adaptive whisker function, the emergence of somatotopically
ordered modular structures may be an inevitable consequence of local
competition for cortical territory driven by input from an array of discrete
sensory organs \citep{purves_iterated_1992}.

\cmnt{The emergence of elongated pre-barrel regions at earlier stages of the
  simulation could be taken as a prediction of the model although where there
  is an absence of the white contours seen in Fig.\,1C/25,000 steps the
  barrels have not yet become well localized and would not be expected to form
  distinct structures.}

@@

\cmnt{It's still a short report if I add a condensed version of supp.tex
  here.}

\cmnt{Software for modelling the cortical sheet as a hexagonal lattice with
  no-flux boundary conditions and an arbitrary boundary shape, for computing
  the flux using a finite difference method, for numerical integration by the
  fourth-order Runge-Kutta method, and for reproducing these results is
  available at \url{https://github.com/ABRG-Models/BarrelEmerge}}

>>>>

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  The
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. \cmnt{The second term on the right
  quantifies the coupling between axon branching and synaptogenesis}. Here $\chi_i=0$ is a placeholder.

@@

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density \cmnt{of unconnected
  branches} at $t=0$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Note that this
  operation is local to individual afferent projections.}

@@

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity \cmnt{and an implicit self/other
  identifier amoungst projections}, and also in terms of cortical space (as
represented by $\chi$). This requirement may account for the absence of
barrels in rodents with larger brains, such as capybara, for which competition
for space is presumably weaker \citep{woolsey_comparative_1975}. Hence,
irrespective of whether barrels are necessary for adaptive whisker function,
the emergence of somatotopically ordered modular structures may be an
inevitable consequence of local competition for cortical territory driven by
input from an array of discrete sensory organs \citep{purves_iterated_1992}.

@@

% Be sure to include/mention:
%
% The no-flux boundary conditions applied in the \cite{karbowski_model_2004}
% model are retained.
%
% Arbitrary boundary shapes were applied

The cortex was modelled as a two dimensional hexagonal lattice which
simplifies the computation of the 2D Laplacian. Within a boundary traced from
rat primary barrel cortex (Fig.\,1A) we set the hex-to-hex distance
$d=0.03$\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

The most involved part of this computation is to find the divergence of the
flux of axonal branching, $\mb{J}_i(\mb{x},t)$, the term in parentheses in
Eq.\,\ref{eq:da}:
%
\begin{equation}
  \label{eq:divJ}
  \nabla\vcdot\mb{J}_i(\mb{x},t) \equiv \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right).
\end{equation}
%
We can define the time-independent sum of the guidance gradients
$\mb{g}_i(\mb{x}) \equiv \sum_{j=1}^{M} \gamma_{i,j} \nabla \rho_j(\mb{x})$
and then apply vector calculus identities to compute
$\nabla\vcdot\mb{J}_i$. Because the divergence operator is distributive,
Eq.~\ref{eq:divJ} can be expanded (dropping references to $\mb{x}$ and $t$ for
clarity):
%
\begin{equation}
\nabla\vcdot\mb{J}_i = \nabla\vcdot\big(D \nabla a_i\big) - \nabla\vcdot\big(a_i \mb{g}_i\big).
\end{equation}
%
Applying the vector calculus product rule identity yields
%
\begin{equation} \label{eq:divJExpanded}
%
\nabla\vcdot\mb{J}_i =
%
D \dvrg a_i % term1
-
a_i\nabla\vcdot\mb{g}_i % term2
-
\mb{g}_i\vcdot\nabla a_i % term3
,
\end{equation}
%
which has three elements to compute: i)~$D \dvrg a_i$ (the Laplacian of
$a_i(\mb{x},t)$); ii)~a time-independent modulator of $a_i(\mb{x},t)$ (because
$\nabla\vcdot\mb{g}_i(\mb{x})$ is a time-independent static field); and iii)~the
scalar product of the static vector field $\mb{g}_i$ and the gradient of
$a_i$. Each of the divergences can be simplified by means of Gauss's Theorem
following \cite{lee_hexagonal_2014}.

(i) The computation of the mean value of the Laplacian across the area of one
hexagon of area $\Omega$ located at position $\mb{p}_0$, with neighbours at
positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \iint_A \dvrg a_i(\mb{x}, t) dxdy & = \frac{1}{\Omega} \oint_c \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big)
\end{split}
\end{equation}
%
where $d$ is the centre-to-centre distance between hexes in the grid, $v =
d/\sqrt{3}$ is the length of each of the 6 edges of the hexagon's perimeter
and $d\gamma$ is an infinitessimally small distance along the perimeter of the
hexagon. The area, $\Omega$, of each hexagon in the lattice is
$\frac{\sqrt{3}}{2}d^2$.

ii) The computation of the second term in Eq.\,\ref{eq:divJExpanded},
$\langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle$, can be written out similarly:

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \iint_A a_i\nabla\vcdot\mb{g}\;dxdy & = \frac{a_i(\mb{p}_0)}{\Omega}  \oint_c \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & \approx \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos \big(\frac{\pi}{3}(j-1)\big) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin \big(\frac{\pi}{3}(j-1)\big) \bigg). \\
%
\end{split}
\end{equation}
%
Both this last expression, and the final expression of Eq.\,\ref{eq:lapl} are
easily computed, comprising short sums of nearest-neighbour relationships.

iii) The final term in Eq.~\ref{eq:divJExpanded} is the scalar product of two
vector fields which is straightforward to compute from their Cartesian
components.

The advantage of separating the computation of Eq.~\ref{eq:divJ} in (i), (ii)
\& (iii) is that we can ensure that the no-flux boundary condition is
fulfilled. On the boundary, the contribution to $\mb{J}$ resulting from the
first term of Eq.~\ref{eq:divJExpanded} can be held at 0 by the ghost cell
method (in which a hex outside the boundary containing the same value as the
hex inside the boundary is imagined such that the flux of $J$ across the
boundary is 0). Then, $\mb{g}_i(\mb{x})$ can be tailored so that it, and its
normal derivative approach 0 at the boundary, ensuring that the second and
third terms of Eq.~\ref{eq:divJExpanded} also contribute nothing to
$\mb{J}$. This is achieved by applying to $\mb{g}(\mb{x})$ a sharp logistic
function of the distance from $\mb{x}$ to the boundary.

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge}. The computations described
in (i), (ii) and (iii) may be found in the code \code{compute\_divJ()} as
\code{term1}, \code{term2} and \code{term3}, respectively.
